An Analysis of Performance Bottlenecks in MRI Pre-Processing

### 2024-05-22 paper/paper.tex
Various edits

\section{Introduction}
% Computational bottleneck
Pre-processing of Magnetic Resonance Imaging (MRI) data is a critical step in neuroimaging. Although established pre-processing pipelines exist, they commonly require expensive computation and produce large output and intermediate data. \HL{These are a well-known challenges to study large subject cohorts in neuroimaging.} Moreover, the execution time of pipelines can hinder clinical applications where timely data analysis is needed. Therefore, the neuroimaging community is constantly seeking ways to improve the performance of MRI pre-processing pipelines.

% HPCs
Researchers broadly adopted High Performance Computing (HPC) to accelerate the MRI processing of large cohorts \HL{leveraging} a large number of CPU cores and high-bandwidth memory to process subjects in parallel. However, \HL{it can be challenging to run HPCs in developing countries due to power restrictions, national network infrastructure, and lack of training and funding}~\cite{NRG-SC23}. \HL{Similarly, researcher use cloud computing, but it requires expertise to setup, can be costly for large studies, and} data transfer can be slow and might be impossible due to ethic concerns related to human data. \HL{In both cases, limitation from the parallel efficiency of the pipeline can lead to under-utilization of the available resources.}

% DL
In recent years, Deep Learning (DL) has shown promising results in MRI pre-preprocessing. For example, tools such as SynthMorph~\cite{Hoffmann2022-hu}, SynthSeg~\cite{Billot2023-vp}, SynthSR~\cite{Iglesias2023-bp}, and SynthStrip~\cite{Hoopes2022-ms} were integrated into FreeSurfer~\cite{Fischl2012-cx} to provide Deep Learning alternatives to traditional algorithms\HL{, providing faster runtime.}
% Moreover, a study~\cite{Pepe2023-dm} showed that the results generated by SynthMorph are numerical more stable than those from FreeSurfer.
However, training a DL model requires large amount of data and specialized hardware, and is computationally expensive. Additionally, \HL{their results} can be challenging to interpret, explaining the slow adoption of DL in MRI pre-processing.

% Alternative solutions
\HL{These techniques require costly hardware to be effective, which can be unacessible to underrepresented researchers.} Alternatively, improved algorithm designs, compression techniques, or reduced- and mixed-precision techniques could provide substantial performance improvements but require substantial effort to implement. Therefore, it is critical to understand the performance bottleneck of MRI pre-processing pipelines to improve their performance.

% - No comprehensive study of the performance bottleneck of MRI pre-processing.
\HL{To the best of our knowledge, this is the first study that} characterize the computational cost of several commonly adopted MRI pre-processing pipelines. The results of our analysis serve as a reference for future efforts to optimize MRI pre-processing pipelines.

% - Focus on MRI pre-processing.
MRI is a standard tool used by neuroscientists to perform clinical diagnosis and for researchers to develop a better understanding of the brain. Three main MRI modalities exist: structural MRI (sMRI), functional MRI (fMRI), and diffusion MRI (dMRI). While other modalities such as electroencephalography (EEG), computed tomography (CT), and positron emission tomography (PET) exist, we focus on MRIs for their broad adoption and non-invasiveness. Moreover, MRI data analysis is challenging due to computationally expensive pre-processing and large output and intermediate data size.

% - Pipelines
Neuroscientists developed various toolboxes to tackle the challenging task of pre-processing MRI data. We focus on the commonly accepted fMRIPrep~\cite{Esteban2019-bl} pipeline as it a comprehensive pre-processing application for both sMRI and fMRI. fMRIPrep combines several pipelines to produce a complete pre-processing pipeline. We also study its sub-pipeline separately to provide a finer grain analysis of the performance bottleneck.

% - VTune profiler
To profile the pipelines, we use the Intel VTune profiler~\cite{vtune-profiler}. The VTune profiler is a multi-language profiler with low performance overhead and provides information of functions at runtime level. We profile the pipeline using \HL{-- single-threaded and 32 threads to simplify the analysis and study the the multi-threading performance, repsectively}. We aggregate subjects' results for each pipeline to obtain an average performance profile. We compiled each pipeline with debug information to obtain human readable information on function and module names in VTune.

\section{Materials \& Methods}
\subsection{Pipelines}

fMRIPrep is a commonly used fMRI and sMRI pre-processing pipeline for neuroimaging, which is built on top of several well-known neuroimaging toolboxes such as ANTs, FSL, FreeSurfer and AFNI. Although the processing steps may vary based on the type of input data, the sMRI pipeline generally uses ANTs BrainExtraction~\cite{Tustison2010-gg,Avants2008-ea} to perform intensity corretion and skull-stripping of a T1 weighted image, FSL FAST~\cite{Zhang2001-hx} to segment the tissues of the extracted brain, and ANTs registrationSyN~\cite{Avants2008-ea} to register the segmented brain. When enabled, FreeSurfer recon-all~\cite{Dale1999-wu} is used for surface reconstruction. We mimic the fMRIPrep specific 3 \HL{phases reconall pre-processing, which (1) performs autorecon1 without skull-stripping, (2) imports an external skull-strip from ANTs brainExtraction, and (3) resumes reconstruction with autorecon2 and autorecon3}. The fMRI pipeline generally performs motion correction with FSL MCFLIRT~\cite{Jenkinson2002-od}, slice timing correction with AFNI 3dtshift~\cite{Cox1996-nk}, and co-registration with FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}.

We profile the entire fMRIPrep pipeline to obtain a coarse view of the pipeline's performance bottleneck. Additionlly, we profile the sub-pipelines of fMRIPrep to characterize the performance at a finer grain. While we used the same sub-pipeline and order as in fMRIPrep, we used the default parameters for most of the sub-pipelines which may vary from the default parameters of fMRIPrep. We omitted some sub-pipelines of fMRIPrep as they were not compatible with our dataset. For example, time-slicing correction was already performed.

\HL{ANTs brainExtraction and registrationSyN provides the user with the option to use a single or double precision version of the pipeline, leveraging the Insight Segmentation and Registration Toolkit (ITK)}~\cite{Yoo2002-ve}. \HL{We profile both version to understand the performance benefits of using reduced precision in the pipelines.}

\subsection{Data}
To ensure that our performance profiles are inclusive of different populations, we used data with a wide range of age and equal distribution of sex. We used the the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}. The anatomical, functional, and diffusion data from 20 health individuals was acquired from two different cohorts. The within-subject cohort has nine participants (mean age=43~yrs, std=7~yrs; 4 females) with two session: eyes open and eyes closed. The replication cohort has eleven participants (mean age=27~yrs, std=5~yrs; 6 females) with only the eyes open session.

@@

First, to obtain human readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image. For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}} \lstlistingname~\ref{lst:vtune_example} illustrates the profiling of a pipeline by mounting the VTune binary to the Apptainer image during execution.

\begin{minipage}{\linewidth}
	\lstinputlisting[
		language={sh},
		caption={Profiling a pipeline within a Apptainer container, using VTune profiler.},
		captionpos=b,
		label=lst:vtune_example,
		numbers=left,
		frame=single,
		basicstyle=\footnotesize,
		numbersep=5pt,
		numberstyle=\tiny\color{gray},
		rulecolor=\color{black},
		tabsize=2,
	]{vtune-example.sh}
\end{minipage}

Last, we profile the pipeline using two threading configuration: single-threaded and 32 threads. The single-threaded approach simplifies the analysis and limits the potential overhead from I/O on the profiling. 
The 32-thread approach uses all available CPU cores from a compute node of our infrastructure to study the multi-threading performance of the pipeline. This is a more realistic usage for the user of the pipeline.

Before profiling each pipeline, we transfer the input data from the shared file system to the compute node. After profiling, we transfer back the output to the shared file system. This limits profiling variability due to I/O contingency.

The data generated by the VTune profiler is written to a shared file system, for administrative reasons. In principle, this could increase the profiling overhead. However, in practice there is no impact since the data generated by the VTune profiler is very small and is not written on the same file system as data produced by the pipelines.

@@

The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load. L1 bound occurs when a demand load stall some CPU cores although the data already reside in L1 cache. Similarly, L2, L3, and DRAM bound occurs when a demand load stall the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively. The store bound occurs when a store operation stall the CPU. Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc}. We use the convention from~\cite{Kukunas2015-jd} for the equations.

@@

\HL{We aggregate the profiling results of each pipeline across all subjects to obtain an average performance profile. The Pareto principle is a power law which states that 80\% of the effects come from 20\% of the causes. We hypothesize that the Pareto principle applies to the performance of MRI pre-processing pipelines, which is consistent with the literature}~\cite{Kukunas2015-jd}. \HL{Therefore, for fine-grained analysis, we only study the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the pipeline. This heuristic allows us to focus our efforts on the performance critical sections of a pipeline. We report the mean and standard deviation of the execution time for those functions.}

\subsection{CPU time distribution across functions was long tailed}
Figure~\ref{fig:long-tail-distribution} shows that 80\% of the total CPU time was spent in less than 1.5\% of the function across all pipelines. While consistent with the Pareto principle, the distribution of the functions' CPU time does not follow a Pareto distribution. Using the \textit{powerlaw} Python package~\cite{Alstott2014-gr} we found that the data fit a Zipf's law with exponent $\alpha=1.46$ (see Figure~\ref{fig:zips-law}). Therefore the CPU time of a function is approximately inversely proportional to its rank. This is consistent with the Pareto's principle, but is more extreme. \HL{In other words, most of the CPU time is spent in very few functions. By consequence, future efforts to optimize this minimal set of functions could result in significant performance improvement.}

\begin{figure}[ht]
	\centering
	\includegraphics[width=\linewidth]{figures/global-longtail.png}
	\caption{Distribution of the functions' CPU time. The left y-axis shows the average total CPU Time spent in a function, while the right y-axis show the cumulative CPU time percentage. The x-axis percentage of functions ordered by decreasing CPU time. The data includes all functions from all pipelines. }
	\label{fig:long-tail-distribution}
\end{figure}

\subsection{Main Bottleneck: Linear Interpolation}
Table~\ref{extab:interpolation} shows that interpolation is a critical part for all pipelines except FreeSurfer recon-all, contributing between 32.20\% to 62.70\% of the total CPU time. Overall, the number of functions using interpolation is low with less than 20 in each pipelines. In other words, few interpolation functions contribute to a large portion of the CPU time. While each interpolation have low computational cost, the large amount of operations make it a significant bottleneck. Therefore, small optimization to the interpolation operation would bring substantial benefits to these pipelines.

\csvnames{csvcol}{1=\pipeline, 2=\nfunc, 3=\cputime}
\csvstyle{interpolation}{
	tabular = |r|c|c|,
	table head = \hline Pipeline & \# of functions & \% CPU Time \\
	& with interpolation & from interpolation\\\hline\hline,
	late after line = \\\hline,
	respect all,
	csvcol
}
\begin{table*}[ht]
	\centering
	\csvreader[interpolation]{tables/interpolation.csv}{}
	{\pipeline & \nfunc & \tablenum[round-precision=2]{\cputime}}
	\caption{Contribution of interpolation to the pipelines' total CPU time. The percentage is the average sum of CPU time of functions using interpolation. The data includes all functions; not only the top 80\% of the CPU time.}
	\label{extab:interpolation}
\end{table*}

Despite an extensive literature on linear interpolation \MD{TODO cite it}, no particular optimization techniques are available in the MRI pre-processing pipelines we profiled. We think it would be interesting to explore some simple optimization techniques. For example, the authors in \cite{Canelhas2018-vs} review different linear interpolation methods ranging from poorly performing nearest-neighbor to classical trilinear interpolation. We think it would be interesting for future work to explore these different alternatives to accelerate interpolation in the pipelines, while considering the trade-off between accuracy and computational cost. Reduced- or mixed-precision techniques could also be explored to accelerate interpolation, with similar trade-off. Optimizing the data format used for interpolation would reduce the computational cost, memory footprint, and storage requirements. Improving the performance of interpolation would have a large impact for 3D MRI imaging processing, and other fields such as climate simulation, astronomical imaging, or 3D image reconstruction.

\subsection{Impact of Memory Bounded Functions}
Table~\ref{extab:memory-bound} shows that fetching data from disk \HL{(Memory bound - sum of other metrics)} is the main source of cycles where the CPU is starved. Following, the L1 bound has the highest cache-level contribution to the memory bound of the pipelines. ANTS pipelines have the highest L1 bound, followed by FreeSurfer recon-all and FSL MCFLIRT. L2 and L3 bound is low for all pipeline, although highest for FreeSurfer recon-all. The DRAM bound is high for both FSL FLIRT and FreeSurfer recon-all, followed by ANTS brainExtraction in single and double precision. Store bound is low for all pipelines. We note an interesting relationship between the low L1 bound and the high DRAM bound of FreeSurfer recon-all and FSL FLIRT, compared to the other pipelines.

Optimizing the data access of the pipeline could reduce the number of data load. This would potentially have a large impact on the performance of the pipelines. Alternatively, the use of reduced precision could reduce the memory footprint, thus the memory bound. To reduce the memory bound impact of fetching data from disk, it could be possible to convert and store the data to a lower precision before pre-processing. In both cases, the impact of reduced precision on the accuracy of the pipelines should be studied.
\MD{TODO TG: You could cite some of Valerie's works in this direction.}

\csvnames{csvcol}{1=\pipeline, 2=\mem, 3=\la, 4=\lb, 5=\lc, 6=\dram, 7=\store}
\csvstyle{memory_bound}{
	tabular = |r|c|c|c|c|c|c|,
	table head = \hline Pipeline &  \% Memory Bound &  \% L1 Bound &  \% L2 Bound &  \% L3 Bound & \% DRAM Bound & \% Store Bound\\\hline\hline,
	late after line = \\\hline,
	respect all,
	csvcol
}
\begin{table*}[ht]
	\centering
	\csvreader[memory_bound]{tables/memory_bound.csv}{}
	{\pipeline & \tablenum[round-precision=2]{\mem} & \tablenum[round-precision=2]{\la} & \tablenum[round-precision=2]{\lb} & \tablenum[round-precision=2]{\lc} & \tablenum[round-precision=2]{\dram} & \tablenum[round-precision=2]{\store}}
	\caption{Impact of data load on stalled CPU cycles. The values are the summation of the metric weighted by each function CPU time. This represent the percentage of the total CPU time stalled by each metrics. We show the average value across all subjects execution with one thread.}
	\label{extab:memory-bound}
\end{table*}

\subsection{ANTs: Single vs. Double Precision}
Figure~\ref{fig:makespan-ants} shows the makespan of ANTS brainExtraction and ANTS registrationSyN in double and single precision. For both pipelines, the makespan is significantly higher in single precision. This is unexpected, as the floating-point arithmetic operations are supposed to be faster in single precision~\cite{Wang2018-jv}.

@@

ANTs registrationSyN is a three stage process (Euler transform, affine transform, and SyN registration), each having 4 level with a specified number of maximum iterations. We focus on the SyN registration stage since it is multiple order longer to excecute. The number of iterations between both version of ANTs registrationSyN is similar. In fact, for the last two levels, the maximum specified number of iterations is reached across all but one execution. However, the single precision version of ANTS registrationSyN took \HL{approximately 20-25\%} longer per iteration than the double precision version for the last registration stage (Figure~\ref{fig:mean-time-per-iteration-ants}). Therefore, the slowdown is not due to the convergence of the algorithm, but rather the execution time for each iteration.

After further analysis of the ANTs registrationSyN single-precision pipeline, we found that the template\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} forces the output of its function to be in double precision, with it's \textit{EvaluateAtContinuousIndex} contributing most of the CPU time. Therefore, the pipeline wastes CPU cycles to convert the input data to double in ITK, and potentially to cast back to float in ANTS. We think this bug causes the slow-down. We tried recompiling a version of ITK to fix this issue, but we were unsuccessful due to the size and complexity of the code base. This example shows that the use of reduced precision is not trivial and require a deep understanding of the code base. Moreover, benchmark should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline. We reported the issue on ITK GitHub repository (see issue \#4593~\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}}).

\begin{figure}
	\includegraphics[width=\linewidth]{figures/ants-registrationSyN-iteration-mean.png}
	\caption{Mean time per iteration for ANTS registrationSyN in double and single precision. Only the SyN Registration stage is shown as the two earlier stage were near zero time. The error bars show the standard deviation.}
	\label{fig:mean-time-per-iteration-ants}
\end{figure}

\subsection{FreeSurfer: Thread-Synchronization}
Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant difference in the pipeline profile between single-threaded and multi-threaded execution. In the multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP is a major bottleneck for the pipeline, accounting for at least 76\% of the pipeline runtime. FreeSurfer uses the multi-processing API provided by OpenMP. Moreover, the y-axis scale is multi folds larger than for the single-threaded executions. This indicates that the multi-threaded execution is significantly slower than the single-threaded execution.

>>>>

\section{Introduction}
% Computational bottleneck
Pre-processing of Magnetic Resonance Imaging (MRI) data is a critical step in neuroimaging. Although established pre-processing pipelines exist, they commonly require extensive amounts of computation and produce large volumes of output and intermediate data. Such requirements create challenges when studying large neuroimaging cohorts, and they limit clinical applicability when timely data analysis is needed. Therefore, the neuroimaging community is constantly seeking for ways to improve the computational performance of MRI pre-processing pipelines. In this paper, we characterize the computational cost of several commonly adopted MRI pre-processing pipelines, providing a reference benchmark for future efforts to optimize MRI pre-processing pipelines.

% % HPCs
% Researchers broadly adopted High Performance Computing (HPC) to accelerate the MRI processing of large cohorts \HL{leveraging} a large number of CPU cores and high-bandwidth memory to process subjects in parallel. However, \HL{it can be challenging to run HPCs in developing countries due to power restrictions, national network infrastructure, and lack of training and funding}~\cite{NRG-SC23}. \HL{Similarly, researcher use cloud computing, but it requires expertise to setup, can be costly for large studies, and} data transfer can be slow and might be impossible due to ethic concerns related to human data. \HL{In both cases, limitation from the parallel efficiency of the pipeline can lead to under-utilization of the available resources.}

% % DL
% In recent years, Deep Learning (DL) has shown promising results in MRI pre-preprocessing. For example, tools such as SynthMorph~\cite{Hoffmann2022-hu}, SynthSeg~\cite{Billot2023-vp}, SynthSR~\cite{Iglesias2023-bp}, and SynthStrip~\cite{Hoopes2022-ms} were integrated into FreeSurfer~\cite{Fischl2012-cx} to provide Deep Learning alternatives to traditional algorithms\HL{, providing faster runtime.}
% % Moreover, a study~\cite{Pepe2023-dm} showed that the results generated by SynthMorph are numerical more stable than those from FreeSurfer.
% However, training a DL model requires large amount of data and specialized hardware, and is computationally expensive. Additionally, \HL{their results} can be challenging to interpret, explaining the slow adoption of DL in MRI pre-processing.

% % Alternative solutions
% \HL{These techniques require costly hardware to be effective, which can be unacessible to underrepresented researchers.} Alternatively, improved algorithm designs, compression techniques, or reduced- and mixed-precision techniques could provide substantial performance improvements but require substantial effort to implement. Therefore, it is critical to understand the performance bottleneck of MRI pre-processing pipelines to improve their performance.

% - No comprehensive study of the performance bottleneck of MRI pre-processing.

% - Focus on MRI pre-processing.
MRI is a standard tool used by neuroscientists to perform clinical diagnosis and for researchers to develop a better understanding of the brain. Three main MRI modalities exist: structural MRI (sMRI), functional MRI (fMRI), and diffusion MRI (dMRI). While other modalities such as electroencephalography (EEG), computed tomography (CT), and positron emission tomography (PET) exist, we focus on MRI for its broad adoption in research and non-invasiveness. Moreover, MRI data analysis is challenging due to computationally expensive pre-processing and large output and intermediate data size.

% - Pipelines
Neuroscientists developed various toolboxes to tackle the challenging task of pre-processing MRI data. We focus on the commonly accepted fMRIPrep~\cite{Esteban2019-bl} pipeline as it a comprehensive pre-processing application for both sMRI and fMRI. fMRIPrep combines several pipelines to produce a complete pre-processing pipeline. We study fMRIPrep's sub-pipelines separately, to provide a finer grained analysis of the performance bottlenecks.

% - VTune profiler
To profile the pipelines, we use the Intel VTune tool~\cite{vtune-profiler}, a multi-language profiler with low performance overhead that provides runtime information at the level of functions. We profile the pipeline in single-threaded mode and using 32 threads to account for parallel executions. We aggregate subjects' results for each pipeline to obtain an average performance profile.

\section{Materials \& Methods}
\subsection{Pipelines}
fMRIPrep is a commonly used fMRI and sMRI pre-processing pipeline for neuroimaging, built on top of several well-known neuroimaging toolboxes such as ANTs, FSL, FreeSurfer and AFNI \TG{expand acronyms and add references for each toolbox}. Although processing steps may vary based on the type of input data, the sMRI pipeline generally uses ANTs BrainExtraction~\cite{Tustison2010-gg,Avants2008-ea} to perform intensity correction and skull-stripping of a T1 weighted image, FSL FAST~\cite{Zhang2001-hx} to segment the tissues of the extracted brain, and ANTs registrationSyN~\cite{Avants2008-ea} to register the segmented brain. When enabled, FreeSurfer recon-all~\cite{Dale1999-wu} is used for surface reconstruction. We mimic the fMRIPrep specific 3 phases reconall pre-processing, \TG{I don't understand this sentence} which (1) performs autorecon1 without skull-stripping, (2) imports an external skull-strip from ANTs brainExtraction, and (3) resumes reconstruction with autorecon2 and autorecon3. The fMRI pipeline generally performs motion correction with FSL MCFLIRT~\cite{Jenkinson2002-od}, slice timing correction with AFNI 3dtshift~\cite{Cox1996-nk}, and co-registration with FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}.

We profiled the entire fMRIPrep pipeline to obtain a coarse view of the pipeline's performance bottleneck. Additionally, to characterize the performance at a finer grain, we profiled the sub-pipelines of fMRIPrep, namely \TG{add list}. While we used the same sub-pipeline and order as in fMRIPrep, we used the default parameters for most of the sub-pipelines which may vary from the default parameters of fMRIPrep. We omitted some sub-pipelines of fMRIPrep as they were not compatible with our dataset. For example, slice-timing correction was already performed in our data.

ANTs brainExtraction and registrationSyN are available in single or double precision, leveraging the Insight Segmentation and Registration Toolkit (ITK)~\cite{Yoo2002-ve}. We profiled both versions to understand the performance benefits of using reduced precision in the pipelines.

\subsection{Data}
To ensure that our performance profiles are inclusive of different populations, we used data with a wide range of age and equal distribution of sex. We used the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}. The anatomical, functional, and diffusion data from 20 health individuals was acquired from two different cohorts. The within-subject cohort has nine participants (mean age=43~yrs, std=7~yrs; 4 females) with two sessions: eyes open and eyes closed. The replication cohort has eleven participants (mean age=27~yrs, std=5~yrs; 6 females) with only the eyes open session. \TG{add information on image resolution}

@@

First, to obtain human-readable information on function and module names in VTune, we re-compiled each pipeline with debug information in a Docker image. For use on HPC systems, we created Apptainer~\cite{Kurtzer2017-bu} images using Docker2Singularity\footnote{\href{https://github.com/singularityhub/docker2singularity}{https://github.com/singularityhub/docker2singularity}}. \lstlistingname~\ref{lst:vtune_example} illustrates the profiling of a pipeline by mounting the VTune binary to the Apptainer image during execution.

\begin{minipage}{\linewidth}
	\lstinputlisting[
		language={sh},
		caption={Profiling a pipeline within an Apptainer container, using VTune profiler.},
		captionpos=b,
		label=lst:vtune_example,
		numbers=left,
		frame=single,
		basicstyle=\footnotesize,
		numbersep=5pt,
		numberstyle=\tiny\color{gray},
		rulecolor=\color{black},
		tabsize=2,
	]{vtune-example.sh}
\end{minipage}

We profiled the pipeline using two threading configuration: single-threaded and 32 threads. The single-threaded approach simplified the analysis and limited the potential I/O overhead on the profiling. 
The 32-thread approach used all available CPU cores from a compute node of our infrastructure to study the multi-threading performance of the pipeline. This configuration provided a more realistic usage of the pipeline.

Before profiling each pipeline, we transferred the input data from the shared file system to the compute node. After profiling, we transferred back the output to the shared file system. This limited profiling variability due to I/O contingency.

The data generated by the VTune profiler was written to a shared file system, to simplify post-processing. In principle, this could have increased the profiling overhead. However, in practice there was no impact since the data generated by the VTune profiler was very small and was not written on the same file system as data produced by the pipelines.

@@

The memory bound metrics counts the number of cycles where the CPU is starved while waiting for in-flight memory demand load. L1 bound occurs when a demand load stalls CPU cores although the data already reside in L1 cache. Similarly, L2, L3, and DRAM bound occurs when a demand load stalls the CPU and there is a cache miss in the previous level of cache, but the data reside in the L2, L3, and DRAM respectively. The store bound occurs when a store operation stalls the CPU. Supplementary Figure~\ref{fig:memory-metrics} shows the monitoring events and equation to derive the metrics defining the memory bound characterization~\cite{Intel2006-lc}. We use the convention from~\cite{Kukunas2015-jd} for the equations.

@@

We aggregate the profiling results of each pipeline across all subjects to obtain an average performance profile. Consistently with~\cite{Kukunas2015-jd}, we hypothesized that the Pareto principle applied to the performance of MRI pre-processing pipelines, we therefore only studied the functions with the largest contribution to CPU time, up to 80\% cumulative total runtime of the pipeline. This heuristic allowed us to focus our efforts on the performance critical sections of a pipeline. We report the mean and standard deviation of the execution time for those functions.

\subsection{CPU time distribution across functions was long-tailed}
Figure~\ref{fig:long-tail-distribution} shows that 80\% of the total CPU time was spent in less than 1.5\% of the function across all pipelines. While consistent with the Pareto principle, the distribution of the functions' CPU time did not follow a Pareto distribution. Using the \textit{powerlaw} Python package~\cite{Alstott2014-gr}, we found that the data fit a Zipf law with exponent $\alpha=1.46$ (Supplementary Figure~\ref{fig:zips-law}). Therefore, the CPU time of a function was approximately inversely proportional to its rank. Thus, future efforts to optimize this minimal set of functions could result in significant performance improvements.

\begin{figure}[ht]
	\centering
	\includegraphics[width=\linewidth]{figures/global-longtail.png}
	\caption{Distribution of the functions' CPU time. The left y-axis shows the average total CPU Time spent in a function, while the right y-axis show the cumulative CPU time percentage. The x-axis is the percentage of functions ordered by decreasing CPU time. The data includes all functions from all pipelines. }
	\label{fig:long-tail-distribution}
\end{figure}

\subsection{Main Bottleneck: Linear Interpolation}
Table~\ref{extab:interpolation} shows that interpolation was a critical part for all pipelines except FreeSurfer recon-all, contributing between 32.20\% to 62.70\% of the total CPU time. Overall, the number of functions using interpolation was low, with less than 20 in each pipeline: few interpolation functions contributed to a large portion of the CPU time. While each interpolation had low computational cost, the large amount of operations resulted in a significant bottleneck. Therefore, optimizing interpolation would bring substantial benefits to these pipelines.

\csvnames{csvcol}{1=\pipeline, 2=\nfunc, 3=\cputime}
\csvstyle{interpolation}{
	tabular = |r|c|c|,
	table head = \hline Pipeline & \# of functions & \% CPU Time \\
	& with interpolation & spent in interpolation\\\hline\hline,
	late after line = \\\hline,
	respect all,
	csvcol
}
\begin{table*}[ht]
	\centering
	\csvreader[interpolation]{tables/interpolation.csv}{}
	{\pipeline & \nfunc & \tablenum[round-precision=2]{\cputime}}
	\caption{Contribution of interpolation to the pipelines' total CPU time. The percentage is the average sum of CPU time of functions using interpolation. The data includes all functions from all pipelines.}
	\label{extab:interpolation}
\end{table*}

\subsection{Impact of Memory Bounded Functions}
Table~\ref{extab:memory-bound} shows that fetching data from disk (Memory bound - sum of other metrics) \TG{I don't understand this subtraction} was the main cause for starving the CPU. Next, the L1 bound had the highest cache-level contribution to the memory bound of the pipelines. ANTS pipelines had the highest L1 bound, followed by FreeSurfer recon-all and FSL MCFLIRT. L2 and L3 bound was low for all pipelines, although higher for FreeSurfer recon-all. The DRAM bound was high for both FSL FLIRT and FreeSurfer recon-all, followed by ANTS brainExtraction in single and double precision. Store bound was low for all pipelines. We note an interesting relationship between the low L1 bound and the high DRAM bound of FreeSurfer recon-all and FSL FLIRT, compared to the other pipelines.

Optimizing the data access of the pipeline could reduce the number of data load. This would potentially have a large impact on the performance of the pipelines. Alternatively, the use of reduced precision could reduce the memory footprint, thus the memory bound. To reduce the memory bound impact of fetching data from disk, it could be possible to convert and store the data to a lower precision before pre-processing. In both cases, the impact of reduced precision on the accuracy of the pipelines should be studied.

\csvnames{csvcol}{1=\pipeline, 2=\mem, 3=\la, 4=\lb, 5=\lc, 6=\dram, 7=\store}
\csvstyle{memory_bound}{
	tabular = |r|c|c|c|c|c|c|,
	table head = \hline Pipeline &  \% Memory Bound &  \% L1 Bound &  \% L2 Bound &  \% L3 Bound & \% DRAM Bound & \% Store Bound\\\hline\hline,
	late after line = \\\hline,
	respect all,
	csvcol
}
\begin{table*}[ht]
	\centering
	\csvreader[memory_bound]{tables/memory_bound.csv}{}
	{\pipeline & \tablenum[round-precision=2]{\mem} & \tablenum[round-precision=2]{\la} & \tablenum[round-precision=2]{\lb} & \tablenum[round-precision=2]{\lc} & \tablenum[round-precision=2]{\dram} & \tablenum[round-precision=2]{\store}}
	\caption{Impact of data load on stalled CPU cycles. The values are the summation of the metric weighted by each function CPU time. This represents the percentage of the total CPU time stalled by each metrics \TG{not sure what this means}. \TG{Memory bound is not the sum of all metrics} We show the average value across all subjects execution with one thread.}
	\label{extab:memory-bound}
\end{table*}

\subsection{ANTs: Single vs. Double Precision}
Figure~\ref{fig:makespan-ants} shows the makespan of ANTS brainExtraction and ANTS registrationSyN in double and single precision. For both pipelines, the makespan was significantly higher in single precision than in double precision. This was unexpected, as the floating-point arithmetic operations are supposed to be faster in single precision~\cite{Wang2018-jv} than in double precision.

@@

ANTs registrationSyN is a three-stage process (Euler transform, affine transform, and SyN registration), where each stage has 4 levels \TG{levels of what?} with a specified number of maximum iterations. We focused on the SyN registration stage since it was substantially longer to execute than the other stages. The number of iterations between both versions of ANTs registrationSyN was similar. In fact, for the last two levels, the maximum specified number of iterations was reached across all but one execution. However, the single precision version of ANTS registrationSyN took approximately 20-25\% longer per iteration than the double precision version for the last registration stage (Figure~\ref{fig:mean-time-per-iteration-ants}). Therefore, the slowdown was not due to the convergence of the algorithm, but rather the execution time for each iteration \TG{add data on this to Figure}.

After further analysis of the ANTs registrationSyN single-precision pipeline, we found that the use of templates in ANTs' C++ code\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/blob/d9c585d96359bf304ad3047148cee81bf27ac0c1/Modules/Core/ImageFunction/include/itkVectorInterpolateImageFunction.h\#L46-L48}{itkVectorInterpolateImageFunction.h}} forces the output of its \TG{clarify what 'its' refers to} function to be in double precision, with \textit{EvaluateAtContinuousIndex} contributing most of the CPU time. Therefore, the pipeline wastes CPU cycles to convert the input data to double-precision in ITK, and potentially to cast back to single-precision in ANTs. We think this bug causes the slow-down. We tried recompiling a version of ITK to fix this issue, but we were unsuccessful due to the size and complexity of the code base. This example shows that the use of reduced precision is not trivial and requires a deep understanding of the code base. Moreover, benchmarks should be performed to ensure that the reduced precision does not impact the performance or accuracy of the pipeline. We reported this issue in the ITK GitHub repository\footnote{\href{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}{https://github.com/InsightSoftwareConsortium/ITK/issues/4593}}.

\begin{figure}
	\includegraphics[width=\linewidth]{figures/ants-registrationSyN-iteration-mean.png}
	\caption{Average time per iteration for ANTS registrationSyN in double and single precision. Only the SyN Registration stage is shown as the two earlier stage were near zero time. The error bars show the standard deviation.}
	\label{fig:mean-time-per-iteration-ants}
\end{figure}

\subsection{FreeSurfer: Thread-Synchronization}
Figure~\ref{fig:hotspots-freesurfer-reconall} shows a significant difference in the pipeline profile between single-threaded and multi-threaded executions. In the multi-threaded executions (Figure~\ref{subfig:hotspots-freesurfer-reconall-32threads}), OMP \TG{expand acronym} was a major bottleneck for the pipeline, accounting for at least 76\% of the pipeline runtime. FreeSurfer used the multi-processing API provided by OpenMP. Moreover, the y-axis scale is multi folds larger than for the single-threaded execution, indicating that the multi-threaded execution was significantly slower than the single-threaded execution.

\begin{figure*}
	\centering
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{Single-threaded}
		\label{subfig:hotspots-freesurfer-reconall-1thread}
		\includegraphics[width=\textwidth]{figures/hotspots-1threads-freesurfer-reconall-simple.png}
	\end{subfigure}
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{Multi-threaded (32 threads)}
		\label{subfig:hotspots-freesurfer-reconall-32threads}
		\includegraphics[width=\textwidth]{figures/hotspots-32threads-freesurfer-reconall-simple.png}
	\end{subfigure}
	\caption{FreeSurfer recon-all analysis. The y-axis shows the average CPU time spent in each function, with error bars showing the standard deviation \TG{across subjects?}. The x-axis shows the function ordered by decreasing CPU time grouped by module. We omitted function names for clarity. The function ID are dependent to each plot. Supplementary materials show the mapping of the function ID to the function name for each plot.}
	\label{fig:hotspots-freesurfer-reconall}
\end{figure*}
